---
title: "Hands-On Exercise 3: Programming Interactive Data Visualization with R"
author: "Bhairavi"
date: "4 February 2023"
---

## Getting Started

The code chunk below installs and launches the following R packages.

```{r}
pacman::p_load(ggiraph, plotly, gganimate, DT, tidyverse, patchwork)
```

## Importing Data

The code chunk below will read the Exam_data.csv file.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Interactive Data Visualization

### ggiraph methods

ggiraph is a htmlwidget and a ggplot2 extension. It allows ggplot graphics to be interactive. This interactivity is made with ggplot geometries that can take in three arguments:

- tooltip: a column of datasets that contain tooltips to be displayed when the mouse if over elements
- onclick: a column of datasets that contain a JavaScript function to be executed when elements are clicked
- data_id: a column of datasets that contain an id to be associated with elements

When used within a R shiny application, elements associated with an id (data_id) can be selected and manipulated on client and server sides.

#### Tooltip Effects

