---
title: "Hands-on Exercise 4: Visualizing Uncertainty"
author: "Bhairavi"
date: "16 February 2023"
---

# Visualizing Uncertainty of Point Estimates

-   A point estimate is a single number, such as a mean
-   Uncertainty is expressed as standard error, confidence interval or credible interval
-   Do not confuse the uncertainty of a point estimate with the variation in the sample

```{r}
pacman::p_load(tidyverse, plotly, crosstalk, DT, ggdist, gganimate)
exam <- read_csv("data/Exam_data.csv")
```

The code chunk below performs the following: - groups the observation by RACE - computes the count of observations, means, standard deviation and standard error of Maths by RACE - saves the output as a tibble data table called my_sum

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(n = n(),
            mean = mean(MATHS),
            sd = sd(MATHS)) %>%
  mutate(se = sd/sqrt(n-1))

knitr::kable(head(my_sum),
             format = "html")
```

The code chunk below is used to reveal the standard error of mean maths score by race.

```{r}
ggplot(my_sum) +
  geom_errorbar(aes(x = RACE,
                    ymin = mean-se,
                    ymax = mean+se),
                width = 0.2,
                colour = "black",
                alpha = 0.9,
                size = 0.5) +
  geom_point(aes(x = RACE,
                 y = mean),
             stat = "identity",
             color = "red",
             size = 1.5,
             alpha = 1) +
  ggtitle("Standard error of mean maths score by race")
```

# ggdist package

-   ggdist is a package that provides a flexible set of ggplot2 geoms and stats designed especially for visualizing distributions and uncertainty
-   designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization
-   for frequentist models, one visualizes confidence distributions or bootstrap distribution
-   for Bayesian models, one visualizes probability distributions

### ggdist methods

In the code chunk below, stat_pointinterval() of ggdist is used to build a visual for displaying distribution of maths score by race.

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(.width = 0.95,
                     .point = median,
                     .interval = qi) +
  labs(title = "Visualizing confidence intervals of mean math score",
       subtitle = "Mean Point + Multiple-interval plot")
```

In the code chunk below, stat_gradientinterval() of ggdist is used to build a visual for displaying distribution of maths score by race.

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(fill = "skyblue",
                        show.legend = TRUE) +
  labs(title = "Visualizing confidence intervals of mean math score",
       subtitle = "Gradient + Interval plot")
```

# Hypothetical Outcome Plots

Step 1: Installing the ungeviz package:

```{r}
devtools::install_github("wilkelab/ungeviz")
```

Step 2: Launch the application in R:

```{r}
library(ungeviz)

ggplot(data = exam,
       aes(x = factor(RACE),
           y = MATHS)) +
  geom_point(position = position_jitter(height = 0.3, width = 0.05),
             size = 0.4,
             color = "#0072B2",
             alpha = 1/2) +
  geom_hpline(data = sampler(25, group = RACE),
              height = 0.6,
              color = "#D55E00") +
  theme_bw() +
  transition_states(.draw, 1, 3)
```